---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## "Reproducible Research: Peer Assessment 1"

#### Preliminary commands
First of all, useful libraries are loaded and the working directory is set.
```{r, echo=TRUE}
library(ggplot2)
library(knitr)
# Set working directory
setwd("~/Documenti/Coursera/DS5_Reproducible_Research/Prj1/RepData_PeerAssessment1/")
```

#### Loading and preprocessing the data

```{r, echo=TRUE}
# Load the dataset
fileN <- "../dataset/activity.csv"
ds_activity <- read.csv(fileN, header=TRUE, sep=",", na.strings="NA")
#
# Convert dates
ds_activity$date <- as.Date(ds_activity$date, "%Y-%m-%d")
str(ds_activity)
```


#### What is mean total number of steps taken per day?
Calculating the sum with a for loop, then mean and median steps per day.

```{r, echo=TRUE}
dates <- data.frame(unique(ds_activity$date))
for (i in 1:dim(dates)[1]) {
   dates[i, 2] <- sum(ds_activity$steps[ds_activity$date == dates[i, 1]], na.rm=FALSE)
}
names(dates) <- c("Day", "Total_steps")
total_mean_steps <- mean(dates[, 2], na.rm=TRUE)
total_median_steps <- median(dates[, 2], na.rm=TRUE)

pl <- ggplot(dates, aes(x = Day, y = Total_steps, group=1)) +
      theme(axis.text.x = element_text(size = rel(1.0), angle = 90, hjust = 1)) +
      labs(x = "Day",
           y = "Total steps") +
      ggtitle("Total steps per day") +
      scale_x_date(breaks="2 day")

pl + geom_bar(stat = "identity", position = "stack", width=0.8)

```

The mean total number of steps taken per day is `r total_mean_steps`, while the median total number of steps taken per day is `r total_median_steps`.
These value do not greatly differ from the ones calculated at the beginning of the assignment. In particular, the mean is the same while the median has slightly changed.
The impact on the estimates of the total daily number of steps is therefore low.


#### What is the average daily activity pattern?

```{r, echo=TRUE}
daily_pattern <- data.frame(unique(ds_activity$interval))
for (i in 1:dim(daily_pattern)[1]) {
   daily_pattern[i, 2] <- 
      mean(ds_activity$steps[ds_activity$interval == daily_pattern[i, 1]], na.rm=TRUE)
}
names(daily_pattern) <- c("Interval", "Average_steps")

max_interval <- daily_pattern[which(daily_pattern$Average_steps==max(daily_pattern$Average_steps)),1]

pl2 <- ggplot(daily_pattern, aes(x = Interval, y = Average_steps, group=1)) +
      theme(axis.text.x = element_text(size = rel(1.0), angle = 90, hjust = 1)) +
      labs(x = "Interval",
           y = "Average steps") +
      ggtitle("Average steps per interval")

pl2 + geom_line(width=0.8)

```

The 5-minutes interval which contains the maximum average steps (averaged over all the days) is `r max_interval` minutes.

#### Imputing missing values
```{r, echo=TRUE}
nas <- as.vector(which(is.na(ds_activity[,1])))
```

The total number of missing values is `r length(nas)`. Now substituting missing values with the average steps for the same interval.

```{r, echo=TRUE}
# New database
ds_activity_2 <- ds_activity

# NA substitution with average steps for the same interval
for (i in 1:length(nas)) {
   nainterv <- ds_activity$interval[nas[i]]
   avg <- daily_pattern$Average_steps[daily_pattern$Interval == nainterv]
   ds_activity_2[nas[i], 1] <- avg
}

dates2 <- data.frame(unique(ds_activity_2$date))
for (i in 1:dim(dates2)[1]) {
   dates2[i, 2] <- 
      sum(ds_activity_2$steps[ds_activity_2$date == dates2[i, 1]], na.rm=FALSE)
}
names(dates2) <- c("Day", "Total_steps")
total_mean_steps2 <- mean(dates2[, 2], na.rm=TRUE)
total_median_steps2 <- median(dates2[, 2], na.rm=TRUE)

pl3 <- ggplot(dates2, aes(x = Day, y = Total_steps, group=1)) +
   theme(axis.text.x = element_text(size = rel(1.0), angle = 90, hjust = 1)) +
   labs(x = "Day",
        y = "Total steps") +
   ggtitle("Total steps per day") +
   scale_x_date(breaks="2 day")

pl3 + geom_bar(stat = "identity", position = "stack", width=0.8)

```


The mean total number of steps taken per day is `r total_mean_steps2` (was `r total_mean_steps`), while the median total number of steps taken per day is `r total_median_steps2` (was `r total_median_steps`).

#### Are there differences in activity patterns between weekdays and weekends?

As can be noted below, during weekday the average steps per interval start to increase after the 500-min interval, while during weekend the average increases later. During the weekend there are slightly higher peaks between 1200 minutes and 1700 minutes.

```{r, echo=TRUE}
wdwe <- vector(mode="character", length=dim(ds_activity_2)[1])
for (i in 1:dim(ds_activity_2)[1]) {
   wd <- weekdays(ds_activity_2[i, 2])
   if (wd == "Saturday" | wd == "Sunday"){
      wdwe[i] <- "Weekend"
   }
   else {
      wdwe[i] <- "Weekday"
   }
}

wdwe <- as.factor(wdwe)

daily_pattern2 <- data.frame(unique(ds_activity_2$interval))
daily_pattern2 <- rbind(daily_pattern2, daily_pattern2)

n4 <- as.numeric(dim(daily_pattern2)[1])
n2 <- n4/2
n3 <- n2+1

for (i in 1:n2) {
   daily_pattern2[i, 2] <-
      mean(ds_activity_2$steps[
         ds_activity_2$interval == daily_pattern2[i, 1] &
         wdwe == "Weekend"
            ], na.rm=TRUE)
   daily_pattern2[i, 3] <- "Weekend"
}

for (i in n3:n4) {
   daily_pattern2[i, 2] <-
   mean(ds_activity_2$steps[
      ds_activity_2$interval == daily_pattern2[i, 1] &
         wdwe == "Weekday"
      ], na.rm=TRUE)
   daily_pattern2[i, 3] <- "Weekday"
}

names(daily_pattern2) <-
   c("Interval", "Average_steps", "wdwe")

pl4 <- ggplot(daily_pattern2, aes(x = Interval, y = Average_steps, group=1)) +
   theme(axis.text.x = element_text(size = rel(1.0), angle = 90, hjust = 1)) +
   labs(x = "Interval",
        y = "Average steps") +
   ggtitle("Average steps per interval") + 
   geom_line(width=0.8) + 
   facet_grid(wdwe ~ .)

pl4
```
